\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-07-18 paper/ms.tex
typos and captions

The growing field of large-scale time domain astronomy requires methods for
probabilistic data analysis that are computationally tractable, even with
large datasets.
Gaussian Processes are a popular class of models used for this purpose but,
since the computational cost scales, in general, as the cube of the number of
data points, their application has been limited to small datasets.
In this paper, we present a novel method for Gaussian Process modeling in
one-dimension where the computational requirements scale linearly with the
size of the dataset.
We demonstrate the method by applying it to simulated and real astronomical
time series datasets.
These demonstrations are examples of probabilistic inference of stellar
rotation periods, asteroseismic oscillation spectra, and transiting planet
parameters.
The method exploits structure in the problem when the covariance function is
expressed as a mixture of complex exponentials, without requiring evenly
spaced observations or uniform noise.
This form of covariance arises naturally when the process is a mixture of
stochastically-driven damped harmonic oscillators~--~providing a physical
motivation for and interpretation of this choice~--~but we also demonstrate
that it can be a useful effective model in some other cases.
We present a mathematical description of the method and compare it to existing
scalable Gaussian Process methods.
The method is fast and interpretable, with a wide range of potential
applications within astronomical data analysis and beyond.
We provide well-tested and documented open-source implementations of this
method in \project{C++}, \project{Python}, and \project{Julia}.

@@

Our main expertise lies in the field of exoplanet discovery and
characterization where GPs have become a model of choice.
We are confident that the method described in the following pages will benefit
this field, but we also expect that there will be applications in other
branches of astrophysics and beyond.
In \sect{examples}, we present applications of the method to research problems
in stellar rotation (\sect{rotation}), asteroseismic analysis (\sect{astero}),
and exoplanet transit fitting (\sect{transit}).
Some readers might consider starting with these section for motivating
examples before delving into the detailed derivations of the earlier sections.

@@

In the following section, we present a method for improving the cubic scaling
for some use cases.
We call our method and its implementations \celerite.\footnote{The name
\celerite\ comes from the French word \foreign{c\'elerit\'e} meaning the speed
of light in a vacuum.
\response{Throughout the paper, when referring to the \celeriteterm\ model, we
typeset \celeriteterm\ in italics and, in reference to the implementation of
this model, we typeset \celerite\ in sans-serif.}}
The \celerite\ method requires using a specific model
for the covariance $k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$ and, although it
has some limitations, we demonstrate in subsequent sections that it can
increase the computational efficiency of many astronomical data analysis
problems.
The main limitation of this method is that it can only be applied to
one-dimensional datasets, where by ``one-dimensional'' we mean that the
 \emph{input coordinates} $\bvec{x}_n$ are scalar, $\bvec{x}_n \equiv
t_n$.
We use $t$ as the input coordinate because one-dimensional GPs are often
applied to time series data but this isn't a real restriction and the
\celerite\ method can be applied to \emph{any} one-dimensional dataset.
Furthermore, the covariance function for the \celerite\ method is
``stationary''.
In other words, $k_\bvec{\alpha}(t_n,\,t_m)$ is only a function of $\tau_{nm}
\equiv |t_n - t_m|$.

@@

To scale GP models to larger datasets, \citet{Rybicki:1995} presented a method
of computing the first term in \eq{gp-likelihood} in $\mathcal{O}(N)$
operations when the covariance function is given by
\begin{eqnarray}\eqlabel{kernel-simple}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})
\end{eqnarray}
where $\{{\sigma_n}^2\}_{n=1}^N$ are the measurement uncertainties,
$\delta_{nm}$ is the Kronecker delta, and $\bvec{\alpha} = (a,\,c)$.
The intuition behind this method is that, for this choice of $k_\bvec{\alpha}$,
the inverse of $K_\bvec{\alpha}$ is tridiagonal and can be computed
with a small number of operations for each data point.
\response{This model has been generalized to arbitrary mixtures of
exponentials \citep[for example,][]{Kelly:2011}
\begin{eqnarray}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J a_j\,\exp(-c_j\,\tau_{nm}) \quad.
\end{eqnarray}
In this case, the inverse is dense but \eq{gp-likelihood} can still be
evaluated with a scaling of $\mathcal{O}(N\,J^2)$, where $J$ is the number of
components in the mixture \citep{Kelly:2014,Ambikasaran:2015}.
In \sect{factor}, we build on previous work \citep{Ambikasaran:2015} derive a
faster method with the same $\mathcal{O}(N\,J^2)$ scaling that can be used
when $k_\bvec{\alpha}(\tau_{nm})$ is positive definite.}

This kernel function can be made even more general by
introducing complex parameters $a_j \to a_j\pm i\,b_j$ and
$c_j \to c_j\pm i\,d_j$.
In this case, the covariance function becomes
\begin{eqnarray}\eqlabel{celerite-kernel-complex}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J &&\left[
    \frac{1}{2}(a_j + i\,b_j)\,\exp\left(-(c_j+i\,d_j)\,\tau_{nm}\right)
        \right. \nonumber\\
    &&+\left.
    \frac{1}{2}(a_j - i\,b_j)\,\exp\left(-(c_j-i\,d_j)\,\tau_{nm}\right)
\right]
\end{eqnarray}
and, for this function, \eq{gp-likelihood} can still be evaluated with
$\mathcal{O}(N\,J^2)$ operations.
The details of this method and a few implementation considerations are
outlined in the following section but we first discuss some
properties of this covariance function.

@@

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=0.95\textwidth]{figures/sho.pdf}
\caption{(left) The power spectrum of a stochastically-driven simple harmonic
    oscillator (\eqalt{sho-psd}) plotted for several values of the quality
    factor $Q$.
    For comparison, the dashed line shows the Lorentzian function from
    \eq{lorentz-psd} with $c_j = \omega_0/(2\,Q) = 1/20$ and normalized so that
    $S(d_j)/S(0) = 100$.
    (middle) The corresponding autocorrelation functions with the same colors.
    \response{(right) Three realizations from each model in the same colors.}
    \figurelabel{sho}}
\end{center}
\end{figure}

@@

For the experiments with small $N \le 2048$, we also compared the log
determinant calculation to the result calculated using the general log
determinant function provided by the \project{NumPy} project
\citep{Van-Der-Walt:2011}, itself built on the \project{LAPACK}
\citep{Anderson:1999} implementation in the Intel Math Kernel
Library\footnote{\url{https://software.intel.com/en-us/intel-mkl}}.
The right-hand panel of \Figure{error} shows a histogram of the fractional
error in the log determinant calculation for each of the 480~experiments with
$N \le 2048$.
Across the board, these results agree within machine precision and the error
only weakly increases with $N$ and $J$.

>>>>

The growing field of large-scale time domain astronomy requires methods for
probabilistic data analysis that are computationally tractable, even with
large datasets.
Gaussian Processes are a popular class of models used for this purpose but,
since the computational cost scales, in general, as the cube of the number of
data points, their application has been limited to small datasets.
In this paper, we present a novel method for Gaussian Process modeling in
one-dimension where the computational requirements scale linearly with the
size of the dataset.
We demonstrate the method by applying it to simulated and real astronomical
time series datasets.
These demonstrations are examples of probabilistic inference of stellar
rotation periods, asteroseismic oscillation spectra, and transiting planet
parameters.
The method exploits structure in the problem when the covariance function is
expressed as a mixture of complex exponentials, without requiring evenly
spaced observations or uniform noise.
This form of covariance arises naturally when the process is a mixture of
stochastically-driven damped harmonic oscillators~--~providing a physical
motivation for and interpretation of this choice~--~but we also demonstrate
that it can be a useful effective model in some other cases.
We present a mathematical description of the method and compare it to existing
scalable Gaussian Process methods.
The method is fast and interpretable, with a range of potential applications
within astronomical data analysis and beyond.
We provide well-tested and documented open-source implementations of this
method in \project{C++}, \project{Python}, and \project{Julia}.

@@

Our main expertise lies in the field of exoplanet discovery and
characterization where GPs have become a model of choice.
We are confident that this method will benefit this field, but we also expect
that there will be applications in other branches of astrophysics and beyond.
In \sect{examples}, we present applications of the method to research problems
in stellar rotation (\sect{rotation}), asteroseismic analysis (\sect{astero}),
and exoplanet transit fitting (\sect{transit}).
Some readers might consider starting with these section for motivating
examples before delving into the detailed derivations of the earlier sections.

@@

In this paper, we present a method for improving the cubic scaling for some
use cases.
We call our method and its implementations \celerite.\footnote{The name
\celerite\ comes from the French word \foreign{c\'elerit\'e} meaning the speed
of light in a vacuum.
\response{Throughout the paper, when referring to the \celeriteterm\ model, we
typeset \celeriteterm\ in italics and, in reference to the implementation of
this model, we typeset \celerite\ in sans-serif.}}
The \celerite\ method requires using a specific model
for the covariance $k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$ and, although it
has some limitations, we demonstrate in subsequent sections that it can
increase the computational efficiency of many astronomical data analysis
problems.
The main limitation of this method is that it can only be applied to
one-dimensional datasets, where by ``one-dimensional'' we mean that the
 \emph{input coordinates} $\bvec{x}_n$ are scalar, $\bvec{x}_n \equiv
t_n$.
We use $t$ as the input coordinate because one-dimensional GPs are often
applied to time series data but this isn't a real restriction and the
\celerite\ method can be applied to \emph{any} one-dimensional dataset.
Furthermore, the covariance function for the \celerite\ method is
``stationary''.
In other words, $k_\bvec{\alpha}(t_n,\,t_m)$ is only a function of $\tau_{nm}
\equiv |t_n - t_m|$.

@@

To scale GP models to larger datasets, \citet{Rybicki:1995} presented a method
of computing the first term in \eq{gp-likelihood} in $\mathcal{O}(N)$
operations when the covariance function is given by
\begin{eqnarray}\eqlabel{kernel-simple}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})
\end{eqnarray}
where $\{{\sigma_n}^2\}_{n=1}^N$ are the measurement uncertainties,
$\delta_{nm}$ is the Kronecker delta, and $\bvec{\alpha} = (a,\,c)$.
The intuition behind this method is that, for this choice of $k_\bvec{\alpha}$,
the inverse of $K_\bvec{\alpha}$ is tridiagonal and can be computed
with a small number of operations for each data point.
\response{This model has been generalized to arbitrary mixtures of
exponentials \citep[for example,][]{Kelly:2011}
\begin{eqnarray}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J a_j\,\exp(-c_j\,\tau_{nm}) \quad.
\end{eqnarray}
In this case, the inverse is dense but \eq{gp-likelihood} can still be
evaluated with a scaling of $\mathcal{O}(N\,J^2)$, where $J$ is the number of
components in the mixture \citep{Kelly:2014,Ambikasaran:2015}.
In \sect{factor}, we build on previous work \citep{Ambikasaran:2015} to derive
a faster method with the same $\mathcal{O}(N\,J^2)$ scaling that can be used
when $k_\bvec{\alpha}(\tau_{nm})$ is positive definite.}

This kernel function can be made even more general by
introducing complex parameters $a_j \to a_j\pm i\,b_j$ and
$c_j \to c_j\pm i\,d_j$.
In this case, the covariance function becomes
\begin{eqnarray}\eqlabel{celerite-kernel-complex}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J &&\left[
    \frac{1}{2}(a_j + i\,b_j)\,\exp\left(-(c_j+i\,d_j)\,\tau_{nm}\right)
        \right. \nonumber\\
    &&+\left.
    \frac{1}{2}(a_j - i\,b_j)\,\exp\left(-(c_j-i\,d_j)\,\tau_{nm}\right)
\right]
\end{eqnarray}
and, for this function, \eq{gp-likelihood} can still be evaluated with
$\mathcal{O}(N\,J^2)$ operations.
The details of this method and a few implementation considerations are
outlined in \sect{factor}, but we first discuss some properties of this
covariance function.

@@

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=0.95\textwidth]{figures/sho.pdf}
\caption{(left) The power spectrum of a stochastically-driven simple harmonic
    oscillator (\eqalt{sho-psd}) plotted for several values of the quality
    factor $Q$.
    For comparison, the dashed line shows the Lorentzian function from
    \eq{lorentz-psd} with $c_j = \omega_0/(2\,Q) = 1/20$ and normalized so that
    $S(d_j)/S(0) = 100$.
    (middle) The corresponding autocorrelation functions with the same colors.
    \response{(right) Three realizations in arbitrary units from each model in
    the same colors.}
    \figurelabel{sho}}
\end{center}
\end{figure}

@@

For the experiments with small $N \le 2048$, we also compared the log
determinant calculation to the result calculated using the general log
determinant function provided by the \project{NumPy} project
\citep{Van-Der-Walt:2011}, itself built on the \project{LAPACK}
\citep{Anderson:1999} implementation in the Intel Math Kernel
Library\footnote{\url{https://software.intel.com/en-us/intel-mkl}}.
The right-hand panel of \Figure{error} shows a histogram of the fractional
error in the log determinant calculation for each of the 480~experiments with
$N \le 2048$.
These results agree to about $10^{-15}$ and the error only weakly increases
with $N$ and $J$.
